Homoicoicity with FAlgebras

### 2016-06-12 paper/paper.tex
final edits

\usepackage{amsthm}
\makeatletter
\def\th@definition{%
  \thm@notefont{}% same as heading font
  \normalfont % body font
}
\makeatother
\theoremstyle{definition}
\newtheorem{problem}{Problem}
\newtheorem{defn}{Definition}

@@

\usepackage{courier}
%\usepackage{pxfonts}
\usepackage{listings}
\lstset{
    %language=Haskell,
    basicstyle=\ttfamily\footnotesize,
    keywordstyle=\bfseries\ttfamily\footnotesize,
    showstringspaces=false,
    %morekeywords={class,data,type,family,instance, where, ghci},
    morekeywords={ghci},
    %basicstyle=\footnotesize,
    literate={->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
             {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
    %keywordstyle=\lst@ifdisplaystyle\color{blue}\fi,
    %commentstyle=\color{gray}
}
\lstMakeShortInline|

@@

@@

%\title{What's the point of homoiconicity?}
%\title{The (floating) point of homoiconicity}
\title{Homoiconicity Without Template Haskell}
%\subtitle{Functional Pearl}

@@

\begin{abstract}
We say a function is \emph{homoiconic} if it can be converted into an isomorphic abstract syntax tree (AST).
This AST can then be manipulated arbitrarily and converted back into a function.
Template Haskell gives a \emph{compile time} form of homoiconicity to Haskell programs;
but in this paper, we demonstrate another form of homoiconicity that relies on polymorphism and GHC's type programming capabilities.
To motivate our technique,
we construct a small library that automatically transforms numerically unstable code into stable code.

@@

%Imagine you've just downloaded and installed a machine learning library for Haskell.
%You open up ghci and import the corresponding modules.
%In the documentation, you see a function called |logLogistic|,
%so you ask ghci its type.
%\begin{lstlisting}
%ghci> :t logLogistic
%logLogistic :: Floating a => a -> a
%\end{lstlisting}
%Now imagine you want to know how the function is implemented.
%\begin{lstlisting}
%ghci> logLogistic var1 :: AST Floating Var
%\end{lstlisting}
%Oh no!
%The |logLogistic| function has a numerical instability in it.

Imagine you're writing a machine learning library in Haskell.
At some point, you'll need to implement the log-logistic function
\begin{equation}
f(x) = \log\left(\frac{1}{1+\exp{(-x)}}\right)
\label{eq:log-logistic}
\end{equation}
This function is used in logistic regression and many families of neural networks.
As a first attempt, you might write:
\begin{lstlisting}
logLogistic1 :: Floating x => x -> x
logLogistic1 x = log(1/(1+exp(-x)))
\end{lstlisting}
Unfortunately, this straightforward implementation is numerically unstable.
When specialized to the |Double| type,
the |exp| function overflows to infinity for inputs greater than (approximately) $709$.
%This overflow causes |logLogistic| to take the |log| of 0, which is undefined.
Therefore, on an input of $-710$,
|logLogisitic| will take the |log| of 0,
returning negative infinity when the true answer is actually very close to $-710$.
Such an inaccurate answer is unacceptable,
so the function must be rewritten.

But like all good programmers, you're lazy.
Your machine learning library will contain hundreds of numeric functions,
all of which will need to be carefully written to avoid numeric instabilities.
What a maintenance nightmare!
Soon you start daydreaming about a function
\begin{lstlisting}
stabilize :: Floating a => (a -> a) -> (a -> a)
\end{lstlisting}
that takes a naively written floating point function as input and outputs a numerically stable version.
Armed with such a function, you could write a stable implementation of the log-logistic function as
\begin{lstlisting}
logLogistic2 :: Floating x => x -> x
logLogistic2 x = stabilize $ log(1/(1+exp(-x)))
\end{lstlisting}
%stopzone

In this paper, we see how to write functions like |stabilize| that perform program transformations.
These functions follow the following pattern.
First, convert the input function into an abstract syntax tree (AST).
Next, manipulate the AST to remove the numerical instabilities.
Finally, convert the AST back into a callable function.
%The key observation is that certain Haskell functions are \emph{homoiconic}.
%That is, we can convert them into isomorphic abstract syntax trees (ASTs).
%The key observation is that certain Haskell functions (which we call \emph{homogeneous} and \emph{heterogeneous}) can be converted into equivalent abstract syntax trees (ASTs).
%The |stabilize| function converts its input function into an AST,
%then manipulate that AST to remove numerical stabilities,
%and finally convert the AST back into a callable function.

The main idea of the |homoiconic| library is that we can associate with every type class an FAlgebra that captures the algebraic structure of that class.

In Section \ref{sec:homogeneous},
%In Section \ref{sec:homogeneous}, we introduce the idea of a \emph{homogeneous} function and show that these functions are homoiconic.
%In Section \ref{sec:stabilize}, we use the construction above to implement several versions of the |stabilize| function.
%In Section \ref{sec:heterogeneous}, we introduce the idea of \emph{heterogeneous} functions and show that they are also homoiconic.
%Finally, we conclude by introducing a Haskell package called {\ttfamily homoiconic} that makes it easy for you to .

There are two intuitions for homogeneous type classes.
First, they have ASTs that contain only a single type variable.
Second, we can construct an FAlgebra for these type classes.

@@

In this section we first define homogeneous type classes and functions.
Then we show that there is an FAlgebra for each homogeneous type class,
and that this FAlgebra makes homogeneous functions homoiconic.

@@

\begin{defn}
We call a function homogeneous if:
\begin{enumerate}
\item
there is exactly one type variable in the signature, which we denote by |a|;
\item
each of the function's parameters is either
\begin{enumerate}
\item
|a|, or
\item
concrete;
\end{enumerate}
\item
the function's return type is |a|; and
\item
the constraints contain exactly one homogeneous type classes applied to |a| and nothing else.
\end{enumerate}
The |logLogistic1| function is homogeneous,
and Figure \ref{code:ghc} shows many more examples taken from the Prelude's numeric hierarchy.
\end{defn}

@@

Using our uniform representation of an AST,
we can make the |AST| type an instance of any homogeneous type class.
As an example, the |Fractional| instance is shown below.
\begin{lstlisting}
instance View Fractional alg
    => Fractional (AST alg a) where
    (/) e1 e2 = Free $ embedSig $ Sig_div e1 e2
    recip e   = Free $ embedSig $ Sig_recip e
    fromRational r
        = Free $ embedSig $ Sig_fromRational r
\end{lstlisting}
For each function in the type class,
we simply embed the function's constructor into the AST by prepending the expression |Free $ embedSig $|.

@@

Now, when we type a numeric expression into ghci without a type signature,
ghci will evaluate the expression like normal.
\begin{lstlisting}
ghci> (1+2)+3
6
\end{lstlisting}
But when we add the appropriate type signature,
the full AST is displayed.
\begin{lstlisting}
ghci> (1+2)+3 :: AST Floating Double
(((fromInteger 1)+(fromInteger 2))+(fromInteger 3))
\end{lstlisting}
Again, the way ghci works, all integer literals automatically have |fromInteger| applied to them,
which is why this function appears in the AST above.
It would be easy to create a pretty printer that does not display |fromInteger| or the excess parentheses,
but this is left as an exercise for the reader.

\subsection{Showing functions}
Reconsider our |logLogistic1| function.
It has type
\begin{lstlisting}
logLogistic1 :: Floating a => a -> a
\end{lstlisting}
What happens if we specialize the |a| parameter to |AST Floating Double|?
We get a peek at the function's internals.
\begin{lstlisting}
ghci> logLogistic1 1 :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (fromInteger 1))))))
\end{lstlisting}
We don't just have to input single numbers;
we can also input full expressions.
These expressions will also get displayed,
intermixed with the original function.
\begin{lstlisting}
ghci> logLogistic1 (1*3+4) :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (((fromInteger 1)*(fromInteger 3))
    +(fromInteger 4)))))))
\end{lstlisting}
And we can even display the resulting AST of nested function calls.
\begin{lstlisting}
ghci> logLogistic1 (logLogistic1 1)
    :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (log ((fromInteger 1)/((fromInteger 1)
    +(exp (negate (fromInteger 1)))))))))))
\end{lstlisting}
It would be nice if the displayed function could show variables,
rather than just numbers.
To accomplish this, we need a variable type.
\begin{lstlisting}
newtype Var = Var String
\end{lstlisting}
Values of type |Var| cannot be plugged directly into the |logLogistic1| function because |Var| is not an instance of |Floating|.
Fortunately, we have a way to make any type an instance of |Floating| via |Floating|'s |AST| type.
More generically, we define the following variables.
\begin{lstlisting}
var1 :: AST alg Var
var1 = Pure "var1"

@@

var3 :: AST alg Var
var3 = Pure "var3"
\end{lstlisting}
And we can use them to recover the exact AST of a function.
\begin{lstlisting}
ghci> logLogistic1 var1 :: AST Floating Var
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate var1)))))
\end{lstlisting}
At this point, you might be tempted to write a |Show| instance for functions.
\begin{lstlisting}
instance
    ( FAlgebra alg
    , Show (Sig alg (Free (Sig alg) Var))
    ) => Show (AST alg Var -> AST alg Var) where
    show f = show (f var1)
\end{lstlisting}
Unfortunately, if you try to use the instance naively
\begin{lstlisting}
ghci> logLogistic1
\end{lstlisting}
then ghci will complain about ambiguous type variables.
GHC's defaulting mechanism is not powerful enough for it to properly specialize the function without a type signature.
So while this construction does give us a convenient way to inspect the contents of a function,
it's not quite powerful enough for a proper |Show| instance.

@@

When we evaluate an AST for an expression,
we get the same result as if we had just evaluated the expression directly.
\begin{lstlisting}
ghci> logLogistic1 10 :: Double
-4.539889921682063e-5

ghci> runAST (logLogistic1 10 :: AST Floating Double)
-4.539889921682063e-5
\end{lstlisting}
We've finally demonstrated that homogeneous functions are in fact homoiconic.

\subsection{Cheating with template Haskell}

For each homogeneous type class,
there are many boilerplate instances that we need to write.
The \homoiconic library provides a template Haskell function |mkFAlgebra| in the |Homogeneous.FAlgebra| module that generates the |FAlgebra|, |Functor|, |Show|, and |View| instances plus all of the pattern synonyms.

@@

\subsection{Pattern matching the AST}

Consider the following simple,
but numerically unstable function.
\begin{lstlisting}
testFunc1 :: Floating a => a -> a
testFunc1 = log (exp a)
\end{lstlisting}
As in our |logLogistic1| function,
|testFunc| gives incorrect answers when |a| is specialized to |Double|.
Inputs greater than (approximately) 710 cause the |exp| function to overflow to infinity;
the log of infinity is still infinity;
so the returned answer is infinity.
But this wrong.
On the real numbers, the $\log$ function is defined to be the inverse of $\exp$,
so the final result should be exactly the same as the input.
We'll write a program transformation that stabilizes the |testFunc1| function,
then we'll demonstrate how the |ViewPatterns| and |PatternSynonyms| language extensions simplify this transformation.
\label{sec:viewpatterns}

@@

We'll start by correcting the second problem using the |ViewPatterns| language extension.
|ViewPatterns| provides a syntactic sugar that lets us pattern match on the return value of a function.\footnote{
    The GHC wiki \cite{viewpatterns} provides details on exactly how the desugaring works.
}
We can use |ViewPatterns| to write polymorphic patterns by using polymorphic functions in the pattern.
We need a polymorphic function for extracting a |Sig alg a| type from a |Sig Floating a| whenever |alg| is a subclass of |Floating|.
Fortunately, we've already defined such a function:
\begin{lstlisting}
unsafeExtractSig :: View alg1 alg2
    => Sig alg2 a -> Sig alg1 a
\end{lstlisting}
Using it in the pattern gives us
\begin{lstlisting}
logexpAST2 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST2
    (Free (unsafeExtractSig -> Sig_log
    (Free (unsafeExtractSig -> Sig_exp a))))
    = a
\end{lstlisting}
This modified program transformation now has the correct polymorphic type signature
(so we've solved problem 2),
but the pattern matches are even uglier
(so we've made problem 1 worse).

@@

pattern AST_exp :: View Floating alg
    => AST alg a -> AST alg a
pattern AST_exp e
    <- Free (unsafeExtractSig -> Sig_exp e)
\end{lstlisting}
Armed with these pattern synonyms,
we can rewrite our function in a much simpler form.
\begin{lstlisting}
logexpAST3 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST3 (AST_log (AST_exp a)) = a
\end{lstlisting}
This simple function works on our the motivating test case:
\begin{lstlisting}
ghci> logexpAST3 $ testFunc1 var1 :: AST Floating Var
var1
\end{lstlisting}
%stopzone
but it fails on this slightly more complex function:
\begin{lstlisting}
testFunc2 :: Floating a => a -> a
testFunc2 a = 1+exp(log a)
\end{lstlisting}
The problem is that we haven't told our program transformations what to do when they don't pattern match!
When pattern matching fails, we want our transformations to recurse into the subexpressions,
applying the transformations there as well.
In the free monad, recursion is handled by a call to |fmap| and the base case is handled by extracting the contents of |Pure|.
This gives our final version of the program transformation.
\begin{lstlisting}
logexpAST4 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST4 (AST_log (AST_exp a)) = a
logexpAST4 (Free f) = Free (fmap logexpAST4 f)
logexpAST4 (Pure a) = Pure a
\end{lstlisting}
Now we correctly recurse into subexpressions.
\begin{lstlisting}
ghci> logexpAST4 $ testFunc2 var1
    :: AST Floating Var
((fromInteger 1)+var1)
\end{lstlisting}
%stopzone
But there's still one last wrinkle.
What if we need to apply the program transformation multiple times?
Consider the following test function.
\begin{lstlisting}
testFunc3 :: Floating a => a -> a
testFunc3 a = 1+log(log(log(exp(exp(exp a)))))
\end{lstlisting}
Applying our program transformation directly doesn't correct the function because the transformation needs to be applied repeatedly.
\begin{lstlisting}
ghci> logexpAST4 $ testFunc3 var1
    :: AST Floating Var
((fromInteger 1)+(log (log (exp (exp var1)))))
\end{lstlisting}
%stopzone
What we want is the \emph{fixed point} of the program transformation.
That is, we want to apply the program transformation repeatedly until it stops changing the AST.
The |fixAST| function below takes as input a transformation and returns the fixed point of that transformation.
\begin{lstlisting}
fixAST ::
    ( Eq (Sig alg (Free (Sig alg) a))
    , Eq a
    ) => (AST alg a -> AST alg a)
      -> (AST alg a -> AST alg a)
fixAST f ast = if ast==ast'
    then ast
    else fixAST f ast'
    where
        ast' = f ast
\end{lstlisting}
Now we can correctly modify the |testFunc3| function.
\begin{lstlisting}
ghci> fixAST logexpAST4 $ testFunc3 var1
    :: AST Floating Var
((fromInteger 1)+var1)
\end{lstlisting}
%stopzone
And that's all there is to writing a program transformation.
More complex transformations just involve more complex pattern matching.

@@

The transformation in in the previous section is relatively simple.
For example, it could have been implemented using GHC's rewrite rules \cite{jones2001playing}.
Rewrite rules are the primary mechanism for optimising source code in GHC,
but they are limited to transformations that involve no computation.
This limitation prevents rewrite rules from implementing a number of optimizations,
one of them being being constant folding.
Constant folding is the process of reducing an expression containing several constants into a single constant.
For example, converting the expression |(1+2)+3| into just the number |6|.
%(Due to this limitation, GHC has special ad-hoc code just for constant folding.)
Our program transformations, however, can implement arbitrary computation and so are powerful enough to handle constant folding.

@@

Let's see an example of constant folding in action.
First define a simple function with two parameters a handful of constants.
\begin{lstlisting}
func :: Num a => a -> a -> a
func x1 x2 = x1*2+(7+2)*x2
\end{lstlisting}
If we call |func| with two constant parameters,
our |foldConstants| function is able to simplify the resulting expression into a single constant.
\begin{lstlisting}
ghci> fixAST foldConstants (func 2 3)
    :: AST Floating Double
(fromInteger 31)
\end{lstlisting}
Alternatively, we can call the function with two variable parameters.
Only the constants in the original function will be folded.
\begin{lstlisting}
ghci> fixAST foldConstants (func var1 var2)
    :: AST Floating Var
((var1*(fromInteger 2))+((fromInteger 9)*var2))
\end{lstlisting}
And finally, we can call the function with one constant and one variable parameter.
\begin{lstlisting}
ghci> fixAST foldConstants (func var1 3)
    :: AST Floating Var
((var1*(fromInteger 2))+(fromInteger 27))
\end{lstlisting}

@@

Now let's see how to stabilize the |logLogistic1| function.
Stabilizing this function is a bit harder because it requires changing the type signature of the function---the constraints on |logLogistic1| require only a |Floating| instance, but the constraints on |logLogistic2| require both a |Floating| and |Ord| instance.
Recall that one of the requirements for a function to be homogeneous is that it have only a single class in the context.
Therefore the stabilized function |logLogistic2| is not homogeneous.

@@

At last we are ready to see our type class for heterogeneous FAlgebras.
\begin{lstlisting}
class FAlgebra alg where
  data Sig alg (t::[Tag]) a

>>>>

\usepackage{amsthm}
\makeatletter
\def\th@definition{%
  \thm@notefont{}% same as heading font
  \normalfont % body font
}
\makeatother
\theoremstyle{definition}
\newtheorem{problem}{Problem}
\newtheorem{defn}{Definition}
\newtheorem{note}{Note}

@@

\usepackage{courier}
%\usepackage{pxfonts}
\usepackage{listings}
\lstset{
    %language=Haskell,
    basicstyle=\ttfamily\footnotesize,
    keywordstyle=\bfseries\ttfamily\footnotesize,
    showstringspaces=false,
    %morekeywords={class,data,type,family,instance, where, ghci},
    morekeywords={ghci},
    %basicstyle=\footnotesize,
    literate={->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
             {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
             {~}{\tiny{$\sim$}}1
    %keywordstyle=\lst@ifdisplaystyle\color{blue}\fi,
    %commentstyle=\color{gray}
}
\lstMakeShortInline|

@@

@inproceedings{sheard2002template,
  title={Template meta-programming for Haskell},
  author={Sheard, Tim and Jones, Simon Peyton},
  booktitle={Proceedings of the 2002 ACM SIGPLAN workshop on Haskell},
  pages={1--16},
  year={2002},
  organization={ACM}
}

@@

%\title{What's the point of homoiconicity?}
%\title{The (floating) point of homoiconicity}
%\title{Homoiconicity Without Template Haskell}
\title{Homoicoicity with FAlgebras}
%\subtitle{Functional Pearl}

@@

\begin{abstract}
We say a function is \emph{homoiconic} if it can be converted into an isomorphic abstract syntax tree (AST).
This AST can then be manipulated arbitrarily and converted back into a function.
Template Haskell gives a limited form of homoiconicity to Haskell programs,
but this paper introduces another form.
The technique relies on an isomorphism between certain Haskell classes and FAlgebras.
The \homoiconic library provides tools for using the technique.
%To make the technique more widely applicable,
%we introduce the Haskell community to the heterogeneous FAlgebra
%polymorphism and GHC's type programming capabilities.
%We also present the \homoiconic library that facilitates working with this technique.
%To motivate our technique,
%we construct a small library that automatically transforms numerically unstable code into stable code.

@@

Template Haskell \cite{sheard2002template} is the standard tool for metaprogramming in Haskell.
%Using quotations, we can convert Haskell code into an abstract syntax tree (AST),
%and using splices, we can insert arbitrary ASTs into our code.
%Template Haskell is enormously useful, but it also has limitations.
In this paper, we introduce a different style of metaprogramming and an accompanying library called \homoiconic.

For example, imagine you're using a library that exports a function
\begin{lstlisting}
logLogistic :: Floating a => a -> a
\end{lstlisting}
There is no way to use Template Haskell to get the abstract syntax tree (AST) of this function.
With \homoiconic, however, we \emph{can} inspect the |logLogistic| function and get its AST.
A ghci session to do this might look like:
\begin{lstlisting}
ghci> logLogistic var1 :: AST Floating Var
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate var1)))))
\end{lstlisting}
In this paper, we'll call any function homoiconic if we can convert it into an AST.
So the |logLogistic| function above is homoiconic.

The \homoiconic library also lets us perform arbitrary program transformations.
A careful look at the output above will reveal that |logLogistic| is numerically unstable.
When specialized to the |Double| type,
the |exp| function overflows to infinity for inputs greater than (approximately) $709$.
%This overflow causes |logLogistic| to take the |log| of 0, which is undefined.
Therefore, on an input of $-710$,
|logLogisitic| will take the |log| of 0,
returning negative infinity when the true answer is actually very close to $-710$.
Using \homoiconic, we can write a function
\begin{lstlisting}
stabilize :: AST Floating a -> AST Floating a
\end{lstlisting}
that converts the AST of an unstable function into a stable function.
We could write a similar transformation in Template Haskell,
but there would be no way for us to get the AST to apply it on!

The key idea of the \homoiconic library is simple:
Certain type classes have FAlgebras that completely capture their structure.
From FAlgebras, we can create ASTs.
And we can use these ASTs to represent expressions generated by the corresponding type classes.
%It's important to note that \homoiconic is much more limited than Template Haskell in other ways.
%In particular, it will only work on certain (but not all!) expressions.
%\footnote{
%FAlgebras are a standard Haskell design pattern for domain specific languages.
%There are many existing references, but Data Types \`a la Carte \cite{swierstra2008} is probably the most famous.
%}
%From these FAlgebras, we can in turn generate ASTs for functions.

The remainder of this paper describes these constructions.
Section \ref{sec:homogeneous} introduces \emph{homogeneous} type classes and functions.
Homogeneous type classes have a particularly simple FAlgebra construction that is a natural extension of the Data Types \`a la Carte approach \cite{swierstra2008}.
Section \ref{sec:stabilize} illustrates how to write program transformations like |stabilize| with a handful of examples operating on numerical functions.
Finally, Section 4 introduces \emph{heterogeneous} type classes.
The naming stems from the intuition that homogeneous classes generate ASTs that contain a single type,
but heterogeneous classes generate ASTs that contain multiple types.
The resulting constructions for the heterogeneous case is significantly more complicated.

@@

In this section we first define homogeneous type classes and functions.
Then we show that there is an FAlgebra for each homogeneous type class,
and that this FAlgebra makes homogeneous functions homoiconic.
All of the constructions discussed here can be found in \homoiconic's |Homogeneous.FAlgebra| module.

@@

\begin{defn}
We call a function homogeneous if:
\begin{enumerate}
\item
there is exactly one type variable in the signature, which we denote by |a|;
\item
each of the function's parameters is either
\begin{enumerate}
\item
|a|, or
\item
concrete;
\end{enumerate}
\item
the function's return type is |a|; and
\item
the constraints contain exactly one homogeneous type classes applied to |a| and nothing else.
\end{enumerate}
The |logLogistic| function is homogeneous,
and Figure \ref{code:ghc} shows many more examples and counterexamples taken from the |Prelude|'s numeric hierarchy.
\end{defn}

@@

Using our uniform representation of an AST,
we can make the |AST| type an instance of any homogeneous type class.
As an example, the |Fractional| instance is shown below.
\begin{lstlisting}
instance
    ( View Fractional alg
    , View Num alg
    ) => Fractional (AST alg a) where
    (/) e1 e2 = Free $ embedSig $ Sig_div e1 e2
    recip e   = Free $ embedSig $ Sig_recip e
    fromRational r
        = Free $ embedSig $ Sig_fromRational r
\end{lstlisting}
For each function in the type class,
we simply embed the function's constructor into the AST by prepending the expression |Free $ embedSig $|.
The constrainst |View num alg| is not used directly in the instance.
It is inherrited from the |Num (AST alg a)| instance,
where it is used directly.
In general, a type class instance for |AST| will require a |View| constraint for every superclass,
not just those parent classes listed in the context.

@@

Now, when we type a numeric expression into ghci without a type signature,
ghci will evaluate the expression like normal.
\begin{lstlisting}
ghci> (1+2)+3
6
\end{lstlisting}
But when we add the appropriate type signature,
the full AST is displayed.
\begin{lstlisting}
ghci> (1+2)+3 :: AST Floating Double
(((fromInteger 1)+(fromInteger 2))+(fromInteger 3))
\end{lstlisting}
Again, the way ghci works, all integer literals automatically have |fromInteger| applied to them,
which is why this function appears in the AST above.
It would be easy to create a pretty printer that does not display |fromInteger| or the excess parentheses,
%but this is left as an exercise for the reader :)
but we that would take us too far afield here.

\subsection{Showing functions}
Reconsider our |logLogistic| function.
It has type
\begin{lstlisting}
logLogistic :: Floating a => a -> a
\end{lstlisting}
What happens if we specialize the |a| parameter to be an |AST|?
We'll get a peek at the function's internals.
\begin{lstlisting}
ghci> logLogistic 1 :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (fromInteger 1))))))
\end{lstlisting}
We don't just have to input single numbers;
we can also input full expressions.
These expressions will also get displayed,
intermixed with the original function.
\begin{lstlisting}
ghci> logLogistic (1*3+4) :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (((fromInteger 1)*(fromInteger 3))
    +(fromInteger 4)))))))
\end{lstlisting}
And we can even display the resulting AST of nested function calls.
\begin{lstlisting}
ghci> logLogistic (logLogistic 1)
    :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (log ((fromInteger 1)/((fromInteger 1)
    +(exp (negate (fromInteger 1)))))))))))
\end{lstlisting}
It would be nice if the displayed function could show variables,
rather than just numbers.
To accomplish this, we need a variable type.
\begin{lstlisting}
newtype Var = Var String
\end{lstlisting}
Values of type |Var| cannot be plugged directly into the |logLogistic| function because |Var| is not an instance of |Floating|.
Fortunately, we have a way to make any type an instance of |Floating| via |Floating|'s |AST| type.
More generically, we define the following variables.
\begin{lstlisting}
var1 :: AST alg Var
var1 = Pure "var1"

@@

var3 :: AST alg Var
var3 = Pure "var3"
\end{lstlisting}
And we can use them to recover the exact AST of a function.
\begin{lstlisting}
ghci> logLogistic var1 :: AST Floating Var
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate var1)))))
\end{lstlisting}
At this point, you might be tempted to write a |Show| instance for functions.
\begin{lstlisting}
instance
    ( FAlgebra alg
    , Show (Sig alg (Free (Sig alg) Var))
    ) => Show (AST alg Var -> AST alg Var) where
    show f = show (f var1)
\end{lstlisting}
Unfortunately, if you try to use the instance naively
\begin{lstlisting}
ghci> logLogistic
\end{lstlisting}
then ghci will complain about ambiguous type variables.
GHC's defaulting mechanism is not powerful enough for it to properly specialize the function without a type signature.
So while this construction does give us a convenient way to inspect the contents of a function,
it's not quite powerful enough for a proper |Show| instance.

@@

When we evaluate an AST for an expression,
we get the same result as if we had just evaluated the expression directly.
\begin{lstlisting}
ghci> logLogistic 10 :: Double
-4.539889921682063e-5

ghci> runAST (logLogistic 10 :: AST Floating Double)
-4.539889921682063e-5
\end{lstlisting}
We've finally demonstrated that homogeneous functions are in fact homoiconic.

\subsection{Boilerplate}

For each homogeneous type class,
there are many boilerplate instances that we need to write.
The \homoiconic library provides a template Haskell function |mkFAlgebra| in the |Homogeneous.FAlgebra| module that generates the |FAlgebra|, |Functor|, |Show|, and |View| instances plus all of the pattern synonyms (discussed in the next section) for a given class.
All of constructions above can be generated by the single invocation
\begin{lstlisting}
mkFAlgebra ''Fractional
\end{lstlisting}

@@

\subsection{Pattern matching the AST}
\label{sec:patternmatch}

Consider the following simple,
but numerically unstable function.
\begin{lstlisting}
testFunc1 :: Floating a => a -> a
testFunc1 = log (exp a)
\end{lstlisting}
As in our |logLogistic| function,
|testFunc| gives incorrect answers when |a| is specialized to |Double|.
Inputs greater than (approximately) 710 cause the |exp| function to overflow to infinity;
the log of infinity is still infinity;
so the returned answer is infinity.
But this wrong.
On the real numbers, the $\log$ function is defined to be the inverse of $\exp$,
so the final result should be exactly the same as the input.
We'll write a program transformation that stabilizes the |testFunc1| function,
then we'll demonstrate how the |ViewPatterns| and |PatternSynonyms| language extensions simplify this transformation.
\label{sec:viewpatterns}

@@

We'll start by correcting the second problem.
The |ViewPatterns| language extension provides syntactic sugar that lets us pattern match on the return value of a function.\footnote{
    The GHC wiki \cite{viewpatterns} provides details on exactly how the desugaring works.
}
We can use |ViewPatterns| to write polymorphic patterns by using polymorphic functions in the pattern.
We need a polymorphic function for extracting a |Sig alg a| type from a |Sig Floating a| whenever |alg| is a subclass of |Floating|.
Fortunately, we've already defined such a function:
\begin{lstlisting}
unsafeExtractSig :: View alg1 alg2
    => Sig alg2 a -> Sig alg1 a
\end{lstlisting}
Using it in the pattern gives us our second program transformation.
\begin{lstlisting}
logexpAST2 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST2
    (Free (unsafeExtractSig -> Sig_log
    (Free (unsafeExtractSig -> Sig_exp a))))
    = a
\end{lstlisting}
This modified transformation has the correct polymorphic type signature
(so we've solved problem 2),
but the pattern matches are even uglier
(so we've made problem 1 worse).

@@

pattern AST_exp :: View Floating alg
    => AST alg a -> AST alg a
pattern AST_exp e
    <- Free (unsafeExtractSig -> Sig_exp e)
\end{lstlisting}
Armed with these pattern synonyms,
we can rewrite our function in a much simpler form.
\begin{lstlisting}
logexpAST3 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST3 (AST_log (AST_exp a)) = a
\end{lstlisting}
This simple function works on our the motivating test case:
\begin{lstlisting}
ghci> logexpAST3 $ testFunc1 var1 :: AST Floating Var
var1
\end{lstlisting}
%stopzone
but it fails on this slightly more complex function:
\begin{lstlisting}
testFunc2 :: Floating a => a -> a
testFunc2 a = 1+exp(log a)
\end{lstlisting}
The problem is that we haven't told our program transformations what to do when they don't pattern match!
When pattern matching fails, we want our transformations to recurse into the subexpressions,
applying the transformations there as well.
In the free monad, recursion is handled by a call to |fmap| and the base case is handled by extracting the contents of |Pure|.

Here is our final version of the program transformation.
\begin{lstlisting}
logexpAST4 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST4 (AST_log (AST_exp a)) = a
logexpAST4 (Free f) = Free (fmap logexpAST4 f)
logexpAST4 (Pure a) = Pure a
\end{lstlisting}
It correctly recurses into subexpressions.
\begin{lstlisting}
ghci> logexpAST4 $ testFunc2 var1
    :: AST Floating Var
((fromInteger 1)+var1)
\end{lstlisting}
%stopzone
But there's still one last wrinkle.
What if we need to apply the program transformation multiple times?
Consider the following test function.
\begin{lstlisting}
testFunc3 :: Floating a => a -> a
testFunc3 a = 1+log(log(log(exp(exp(exp a)))))
\end{lstlisting}
Applying our program transformation directly doesn't correct the function because the transformation needs to be applied repeatedly.
\begin{lstlisting}
ghci> logexpAST4 $ testFunc3 var1
    :: AST Floating Var
((fromInteger 1)+(log (log (exp (exp var1)))))
\end{lstlisting}
%stopzone
What we want is the \emph{fixed point} of the program transformation.
That is, we want to apply the program transformation repeatedly until it stops changing the AST.
The |fixAST| function below takes as input a transformation and returns the fixed point of that transformation.
\begin{lstlisting}
fixAST ::
    ( Eq (Sig alg (Free (Sig alg) a))
    , Eq a
    ) => (AST alg a -> AST alg a)
      -> (AST alg a -> AST alg a)
fixAST f ast = if ast==ast'
    then ast
    else fixAST f ast'
    where
        ast' = f ast
\end{lstlisting}
Now we can correctly modify the |testFunc3| function.
\begin{lstlisting}
ghci> fixAST logexpAST4 $ testFunc3 var1
    :: AST Floating Var
((fromInteger 1)+var1)
\end{lstlisting}
%stopzone
And that's all there is to writing a program transformation.
More complex transformations just involve more complex pattern matching.

@@

The transformation in in the previous section is relatively simple.
For example, it could have been implemented using GHC's rewrite rules \cite{jones2001playing}.
Rewrite rules are the primary mechanism for optimising source code in GHC,
but they are limited to transformations that involve no computation.
This limitation prevents rewrite rules from implementing a number of optimizations,
one of them being being constant folding.
Constant folding is the process of reducing an expression containing several constants into a single constant.
For example, converting the expression |(1+2)+3| into just the number |6|.
%(Due to this limitation, GHC has special ad-hoc code just for constant folding.)
Our program transformations can implement arbitrary computation and so are powerful enough to handle constant folding.

@@

Let's see an example of constant folding in action.
First define a simple function with two parameters and a handful of constants.
\begin{lstlisting}
func :: Num a => a -> a -> a
func x1 x2 = x1*2+(7+2)*x2
\end{lstlisting}
If we call |func| with two constant parameters,
our |foldConstants| function is able to simplify the resulting expression into a single constant.
\begin{lstlisting}
ghci> fixAST foldConstants (func 2 3)
    :: AST Floating Double
(fromInteger 31)
\end{lstlisting}
Alternatively, we can call the function with two variable parameters.
Only the constants in the original function will be folded.
\begin{lstlisting}
ghci> fixAST foldConstants (func var1 var2)
    :: AST Floating Var
((var1*(fromInteger 2))+((fromInteger 9)*var2))
\end{lstlisting}
And finally, we can call the function with one constant and one variable parameter.
\begin{lstlisting}
ghci> fixAST foldConstants (func var1 3)
    :: AST Floating Var
((var1*(fromInteger 2))+(fromInteger 27))
\end{lstlisting}

@@

Now let's see how to stabilize the |logLogistic| function.
Here's an example of what a stable implementation might look like.
\begin{lstlisting}
logLogistic2 :: (Floating x, Ord x) => x -> x
logLogistic2 x = m+log(1/(exp(m)+exp(-x+m)))
    where
        m = min 0 x
\end{lstlisting}
Notice that it has a different type signature---we've had to add the |Ord| constraint due to the use of the |min| function.
Recall that one of the requirements for a function to be homogeneous is that it have only a single class in the context.
Therefore the stabilized function |logLogistic2| is not homogeneous.

@@

At last we are ready to see our type class for heterogeneous FAlgebras.
\newpage
\begin{lstlisting}
class FAlgebra alg where
  data Sig alg (t::[Tag]) a
